---
title: 'BIOS 640 Week 2: Preparing NHANES Blood Pressure and Demographic Data'
author: "Urvashi"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
  html_document:
    df_print: paged
geometry: margin=0.9in
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      comment = "#>")

# pacman loads (and installs, if needed) everything else, so it has to be
# present first or the report will not knit on another machine.
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(
    tidyverse, # dplyr, tidyr, stringr, purrr, ...
    rio,       # import() / export()
    here,      # project-relative file paths
    janitor,   # clean_names()
    skimr,     # skim()
    knitr      # kable()
)
```

# Introduction

The National Health and Nutrition Examination Survey (NHANES) is a program of
studies that assesses the health and nutritional status of adults and children
in the United States, combining interviews with physical examinations and
laboratory data. This report documents how I prepared **blood pressure** and
**demographic** data from three survey cycles (waves): 2013-2014, 2015-2016 and
2017-2018.

The workflow follows the sections below: *Data import and exploration*,
*Merging*, *Cleaning and recoding*, and *Filtering and reshaping*. Names use
`snake_case` and paths use `here::here()`.

# Data import and exploration

Each wave contributes two files, a demographic file and a blood pressure file,
so six files are imported in total. Names follow a `type_year` pattern so that
any dataset can be identified from its name alone.

```{r import-data}
# Import datasets directly from local project directories using the specified naming convention
demo_2013 <- import(here("data", "raw", "2013-2014_DEMO_H.xpt.txt"), format = "xpt")
demo_2015 <- import(here("data", "raw", "2015-2016_DEMO_I.xpt.txt"), format = "xpt")
demo_2017 <- import(here("data", "raw", "2017-2018_DEMO_J.xpt.txt"), format = "xpt")

bp_2013   <- import(here("data", "raw", "2013-2014_BPX_H.xpt.txt"), format = "xpt")
bp_2015   <- import(here("data", "raw", "2015-2016_BPX_I.xpt.txt"), format = "xpt")
bp_2017   <- import(here("data", "raw", "2017-2018_BPX_J.xpt.txt"), format = "xpt")
```

## Dimensions and variable types

```{r raw-dimensions}
raw_datasets <- list(
  demo_2013 = demo_2013, demo_2015 = demo_2015, demo_2017 = demo_2017,
  bp_2013   = bp_2013,   bp_2015   = bp_2015,   bp_2017   = bp_2017
)

tibble(
  dataset = names(raw_datasets),
  rows    = map_int(raw_datasets, nrow),
  columns = map_int(raw_datasets, ncol),
  types   = map_chr(raw_datasets,
                    \(d) paste(sort(unique(map_chr(d, \(x) class(x)[1]))),
                               collapse = ", "))
) %>% kable(caption = "Dimensions and variable types of the six raw datasets")
```

Every variable in every file arrives as `numeric`. This is a property of the
`.XPT` transport format, which stores only numbers and character strings and
has no concept of a factor, so all of NHANES' categorical codes import as plain
numbers.

## Variables stored in an inappropriate format

Two examples where the imported type does not match what the variable actually
represents:

- **`SEQN` (participant ID):** stored as numeric, but it is an identifier
  rather than a quantity. Its mean is meaningless and nothing is gained by
  ordering participants by it. A `character` type is more suitable, because it
  makes accidental arithmetic impossible.
- **`RIAGENDR` (gender):** stored as numeric codes (1 = male, 2 = female) for a
  nominal category. Treating it as a number invites a nonsensical "mean gender"
  of 1.5. A labelled `factor` is more suitable, because it carries the meaning
  of each code and restricts the variable to its valid values. The same
  argument applies to the ethnicity, education, marital status, family income
  and family size variables.

# Merging the datasets

I kept only the required columns, stacked the three waves with a `wave_id`
column, and joined the two combined datasets. A **full join** keeps every
participant who appears in at least one of the two combined datasets.

```{r merge}
bp_vars <- c("SEQN", "PEASCCT1",
             paste0(rep(c("BPXSY", "BPXDI"), 4), rep(1:4, each = 2)))
demo_vars <- c("SEQN", "RIAGENDR", "RIDAGEYR", "RIDRETH1", "RIDRETH3",
               "DMDEDUC2", "DMDMARTL", "DMDFMSIZ", "INDFMIN2")

# One helper used for both file types: select the required columns from each
# wave, then stack them, letting bind_rows() write the wave label itself.
stack_waves <- function(datasets, vars) {
  datasets %>% map(\(d) select(d, all_of(vars))) %>% bind_rows(.id = "wave_id")
}

bp_all   <- stack_waves(list("2013-2014" = bp_2013, "2015-2016" = bp_2015,
                             "2017-2018" = bp_2017), bp_vars)
demo_all <- stack_waves(list("2013-2014" = demo_2013, "2015-2016" = demo_2015,
                             "2017-2018" = demo_2017), demo_vars)

# Joining on wave_id as well as SEQN keeps a single wave_id column in the
# result, instead of wave_id.x and wave_id.y.
nhanes_merged <- full_join(bp_all, demo_all, by = c("SEQN", "wave_id"))
```

```{r compare-dimensions}
n_both      <- nrow(inner_join(bp_all, demo_all, by = c("SEQN", "wave_id")))
n_only_bp   <- nrow(anti_join(bp_all, demo_all, by = c("SEQN", "wave_id")))
n_only_demo <- nrow(anti_join(demo_all, bp_all, by = c("SEQN", "wave_id")))

tibble(dataset = c("bp_all", "demo_all", "nhanes_merged"),
       rows    = c(nrow(bp_all), nrow(demo_all), nrow(nhanes_merged)),
       columns = c(ncol(bp_all), ncol(demo_all), ncol(nhanes_merged))) %>%
  kable(caption = "Combined datasets versus the merged dataset")
```

The merged dataset has **`r nrow(nhanes_merged)` rows and
`r ncol(nhanes_merged)` columns**.

**Rows.** The merged row count is not the sum of the two inputs. It equals the
row count of `demo_all` exactly, because `r n_only_bp` participants appear in
the blood pressure data without appearing in the demographic data. That zero is
the informative result: every person with a blood pressure reading also has a
demographic record, with no exceptions. This follows from how NHANES is run.
All participants are interviewed, which produces the demographic file, and only
a subset then attends the mobile examination centre where blood pressure is
measured. The blood pressure file is therefore a strict subset of the
demographic file. The remaining `r n_only_demo` participants were interviewed
but never examined, so they enter the merged data with `NA` in every blood
pressure column.

**Columns.** The merged column count is `r ncol(bp_all)` + `r ncol(demo_all)` -
2 = `r ncol(nhanes_merged)`. Two columns are subtracted rather than one because
both `SEQN` and `wave_id` were used as join keys, so each appears once in the
result instead of twice.

# Cleaning and recoding

## Step 1: Review the merged data

```{r skim-raw}
skim_without_charts(nhanes_merged) %>%
  yank("numeric") %>%
  select(skim_variable, n_missing, complete_rate, mean, p0, p100) %>%
  kable(digits = 3)
```
```{r diastolic-zeros}
nhanes_merged %>%
  summarise(across(matches("^BPXDI[1-4]$"), \(x) sum(x == 0, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "reading", values_to = "readings_of_zero") %>%
  kable(caption = "Diastolic readings recorded as zero")
```

Four observations from this summary.

**Every variable except `wave_id` is numeric**, including the participant ID and
every categorical code. `wave_id` is character because it was created from the
names of the list passed to `bind_rows()`.

**Missingness in the blood pressure readings is not uniform.** The first three
systolic and diastolic readings have broadly similar completeness, and the
fourth is missing for the overwhelming majority of participants. This is by
design rather than a data quality problem: a fourth reading is taken only when
one of the earlier readings was interrupted or incomplete.

**`DMDEDUC2` and `DMDMARTL` are missing for participants under 20.** The
codebook labels `DMDEDUC2` as "Education level - Adults 20+", and both questions
are asked only of adults, so the missingness is structural and should not be
treated as data loss.

**Several variables carry reserved codes that are not real values.** `DMDEDUC2`
uses 7 for "refused" and 9 for "don't know", while `DMDMARTL` and `INDFMIN2` use
77 and 99 for the same two answers. Left in place, a "don't know" income would
enter any calculation as the number 99. These have to become `NA`. `RIAGENDR`,
`RIDRETH1`, `RIDRETH3` and `DMDFMSIZ` have no such codes.

**The minimum diastolic reading is zero.** All four diastolic columns have a `p0`
of 0, which is not a physiologically meaningful blood pressure. The NHANES
documentation states, under Data Processing and Editing, that diastolic blood
pressure can be zero, but gives no reason for it. The zeros are therefore
deliberate and documented rather than an import error.

One further quirk worth noting for later: `RIDAGEYR` is **topcoded**, so every
participant aged 80 or older is recorded as exactly 80.

## Step 2: Clean the column names

`clean_names()` standardises everything to `snake_case`. That alone still leaves
names like `riagendr` and `bpxsy1`, which are meaningless to a reader, so the
codebook names are then renamed to describe what they hold.

```{r clean-names, results='hide'}
nhanes_clean <- nhanes_merged %>%
  clean_names() %>%
  rename(participant_id = seqn, gender = riagendr, age_years = ridageyr,
         race_ethnicity_5cat = ridreth1, race_ethnicity_6cat = ridreth3,
         education = dmdeduc2, marital_status = dmdmartl,
         family_size = dmdfmsiz, family_income = indfmin2,
         bp_comment = peascct1) %>%
  rename_with(\(x) str_replace(x, "^bpxsy(\\d)$", "sbp_\\1"), starts_with("bpxsy")) %>%
  rename_with(\(x) str_replace(x, "^bpxdi(\\d)$", "dbp_\\1"), starts_with("bpxdi"))
```

## Step 3: Recode the variables

Each categorical variable becomes a labelled factor using its NHANES codebook.
Listing only the valid codes in `levels` means the reserved codes identified in
Step 1 are converted to `NA` automatically, with no separate filtering step.

Seven variables are recoded (gender, two ethnicity variables, education,
marital status, family income and family size) and an age grouping is derived.
Two representative examples are shown; the full chunk is in the `.Rmd`.

```{r recode-example, eval=FALSE}
gender    = factor(gender, levels = c(1, 2), labels = c("Male", "Female"))
age_group = cut(age_years, breaks = c(0, 18, 30, 45, 60, 75, Inf),
                labels = c("0-17", "18-29", "30-44", "45-59", "60-74", "75+"),
                right = FALSE)
```

```{r recode-full, echo=FALSE}
nhanes_clean <- nhanes_clean %>%
  mutate(
    participant_id = as.character(participant_id),
    wave_id        = factor(wave_id),
    gender = factor(gender, levels = c(1, 2), labels = c("Male", "Female")),
    # RIDRETH1 and RIDRETH3 use DIFFERENT codes for different categories:
    # 5 is "other race" in RIDRETH1 but does not exist in RIDRETH3, where
    # 6 is Non-Hispanic Asian and 7 is "other race".
    race_ethnicity_5cat = factor(
      race_ethnicity_5cat, levels = c(1, 2, 3, 4, 5),
      labels = c("Mexican American", "Other Hispanic", "Non-Hispanic White",
                 "Non-Hispanic Black", "Other race (incl. multiracial)")),
    race_ethnicity_6cat = factor(
      race_ethnicity_6cat, levels = c(1, 2, 3, 4, 6, 7),
      labels = c("Mexican American", "Other Hispanic", "Non-Hispanic White",
                 "Non-Hispanic Black", "Non-Hispanic Asian",
                 "Other race (incl. multiracial)")),
    # Ordered, because the categories have a natural ranking.
    education = factor(
      education, levels = c(1, 2, 3, 4, 5), ordered = TRUE,
      labels = c("Less than 9th grade", "9th-11th grade",
                 "High school graduate/GED", "Some college/AA degree",
                 "College graduate or above")),
    # Unordered: no category is "more" than another.
    marital_status = factor(
      marital_status, levels = c(1, 2, 3, 4, 5, 6),
      labels = c("Married", "Widowed", "Divorced", "Separated",
                 "Never married", "Living with partner")),
    # Code 11 is unused. Codes 12 and 13 are coarse brackets given by
    # respondents who would not name an exact band, so they overlap codes
    # 1-5 and the variable is left unordered.
    family_income = factor(
      family_income, levels = c(1:10, 12, 13, 14, 15),
      labels = c("$0-4,999", "$5,000-9,999", "$10,000-14,999", "$15,000-19,999",
                 "$20,000-24,999", "$25,000-34,999", "$35,000-44,999",
                 "$45,000-54,999", "$55,000-64,999", "$65,000-74,999",
                 "$20,000 and over (unspecified)", "Under $20,000 (unspecified)",
                 "$75,000-99,999", "$100,000 and over")),
    family_size = factor(
      family_size, levels = 1:7, ordered = TRUE,
      labels = c("1 person", "2 people", "3 people", "4 people", "5 people",
                 "6 people", "7 or more people")),
    # right = FALSE makes each band closed on the left, so "18-29" is 18 up to
    # but not including 30. The top band is open because RIDAGEYR is topcoded
    # at 80, so 80 means "80 or older" rather than exactly 80.
    age_group = cut(age_years, breaks = c(0, 18, 30, 45, 60, 75, Inf),
                    labels = c("0-17", "18-29", "30-44", "45-59", "60-74", "75+"),
                    right = FALSE)
  )
```

The age bands follow recognisable life stages rather than equal-width
intervals: minors, young adults, early middle age, late middle age, early old
age, and the topcoded group.

```{r recode-check}
nhanes_clean %>%
  count(age_group, gender) %>%
  pivot_wider(names_from = gender, values_from = n) %>%
  kable(caption = "Participants by age group and gender after recoding")
```

```{r education-check}
nhanes_clean %>%
  count(education) %>%
  kable(caption = "Education level after recoding, including reserved codes now missing")
```

Table 4 shows both variables carrying labels rather than the codes 1 and
table 5 shows the reserved codes 7 and 9 have become `NA` rather than being
counted as if they were education levels.

## Step 4: Duplicates, invalid readings and export

```{r step4}
n_duplicates <- sum(duplicated(nhanes_clean))
nhanes_clean <- distinct(nhanes_clean)
```

`r n_duplicates` duplicate rows were found. This is the expected result: the
join key was the pair `SEQN` and `wave_id`, and NHANES issues each participant
a unique `SEQN` within a wave, so every row is already a distinct
participant-wave combination.

`bp_comment` (`PEASCCT1` in the codebook) is populated **only** when something
prevented the blood pressure examination from being completed normally, and is
missing for everyone whose examination went ahead as intended. The specific
codes differ between waves: 2013-2014 uses a detailed twelve-code scheme, while
the two later waves use four broader categories, and the same number does not
always mean the same thing across them.

Testing for a missing value rather than for particular code numbers therefore
identifies flagged examinations correctly in all three waves without depending
on any wave's coding scheme.

```{r invalid-readings}
nhanes_clean <- nhanes_clean %>%
  mutate(bp_invalid = !is.na(bp_comment)) %>%
  mutate(across(matches("^(sbp|dbp)_[1-4]$"),
                \(x) if_else(bp_invalid, NA_real_, x)))

n_flagged <- sum(nhanes_clean$bp_invalid)
```

`r n_flagged` examinations were flagged and had their readings set to missing.

```{r export}
dir.create(here("data", "clean"), recursive = TRUE, showWarnings = FALSE)
export(nhanes_clean, here("data", "clean", "nhanes_bp_demographics_clean.rds"))
```

The cleaned dataset is exported as `.rds` rather than `.csv` because `.rds`
preserves factor levels, orderings and types, so a later analysis can import it
and start work immediately rather than repeating this whole pipeline.

As a final check, the first rows of the cleaned dataset:

```{r final-check}
nhanes_clean %>%
  select(participant_id, wave_id, gender, age_group, education, sbp_1, dbp_1) %>%
  head(5) %>%
  kable(caption = "First five rows of the cleaned dataset")
```

# Filtering and reshaping

I kept adults with a non-missing age of 21 or older, along with age,
participant ID, gender and the eight blood pressure measurements.

```{r filter-adults}
nhanes_adults <- nhanes_clean %>%
  filter(!is.na(age_years), age_years >= 21) %>%
  select(age_years, participant_id, gender, matches("^(sbp|dbp)_[1-4]$"))
```

The result is in **wide format**. Each participant occupies exactly one row, and
their repeated measurements are spread sideways across eight separate columns
(`sbp_1` to `sbp_4` and `dbp_1` to `dbp_4`). The reading number is encoded in
the column names rather than stored as data, which is the defining
characteristic of wide format.

`pivot_longer()` converts it to **long format**, giving one row per participant
per reading. Each of the eight column names encodes two separate facts, the
measure and the reading number, so `names_to = c(".value", "reading_number")`
splits them: `.value` sends `sbp` and `dbp` to their own columns, while the
trailing digit becomes a `reading_number` variable.

```{r reshape}
nhanes_adults_long <- nhanes_adults %>%
  pivot_longer(cols = matches("^(sbp|dbp)_[1-4]$"),
               names_to = c(".value", "reading_number"), names_sep = "_")

c(wide_rows = nrow(nhanes_adults), long_rows = nrow(nhanes_adults_long))
head(nhanes_adults_long, 4)
```

The long dataset has four times as many rows as the wide one, one per reading,
and the eight measurement columns collapse into `reading_number`, `sbp` and
`dbp`.
